An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess

### 2022-12-04 Manuscript/manuscript.tex
update intro

Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen, and “alleged cheating” in chess. The scandal started when Carlsen, the then reining five-time World Chess Champion, withdrew from the Sinquefield Cup after a stunning upset by then nineteen-year-old Niemann in the tournament’s third round. Carlsen didn’t immediately state the reason as to why he withdrew at first, though many correctly assumed he was holding his tongue in making any sort of public accusation of cheating to avoid any potential retaliatory lawsuit, but it was too late. The public was already running wild with the story, and every armature chess player and data scientist was looking for any indication of foul play given the high-stakes nature of the situation and the people involved.

The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online. Cheating in online chess is very easy. One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next \citep{kenregan2010}. Cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them.

The specific allegations related to how Niemann cheated OTB are unknown. Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game. But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career? Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different. Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games \citep{erikallebest2022}. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs. "difficult" moves.

Item 2 requires an assessment of move quality, which is usually done in centipawns. The centipawn is a unit of measure used in chess to quantify advantage. A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions . Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine. Therefore, average centipawn loss would be a calculation of the average loss incurred by each move over any particular game.

Measures of average centipawn loss can be used to evaluate a player's strength and, theoretically, one can compare ACL to one's ELO rating, a measure of one's relative skill level in zero-sum games such as chess, to observe how well they played with respect to this ranking. The contribution of this paper is to provide an analysis of this method, as a means to detect cheating in chess, by observing suspicious historic activity (i.e., an individual playing well above their rating in terms of ACL) with the goal of scrutinizing whether it should be treated as a "smoking gun" in terms of evidence of cheating; as was the case with Hans Niemann.

The rest of the paper is organized as follows. Section 2 provides specifics about the data used for the ACL versus ELO rating analysis, such as how it was collected and what it does to help answer my research question. Section 3 will provide insight into my methods and computer programs used to perform the analysis. Section 4 concludes with a discussion.

>>>>

Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen,  and “alleged cheating” in the world of professional chess. The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round. This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers.

Carlsen also didn't immediately state the reason as to why he withdrew at first, though many correctly assumed he was holding his tongue in making any sort of public accusation of cheating to avoid a retaliatory lawsuit by Niemann while he didn't have proof. But it was too late, the public was already running wild with the story and amateur and professional entities alike were launching their own investigations on the matter, looking for any indication of foul play.

Although the question of whether Hans Niemann “cheated” originally stemmed from just the one game, it would be nearly impossible to prove foul play by simply analyzing moves without any form of “smoking gun” in terms of evidence, such as the discovery of a concealed communication device or computer on his person. This is primarily due to the modality in which the game was played.

The game between Carlsen and Niemann was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online. As one might guess cheating in online chess is quite easy. One simply has to rely on outside input, commonly through a chess engine (such as Stockfish, AlphaZero, and MuZero) and execute the moves from their game into the engine as they played, following up with whichever move that was suggested to play next. This has become an ever increasing problem as the superiority of top chess engines have been acknowledged ever since IBM's Deep Blue defeated former World Chess Champion Garry Kasparov in 1997.

But, cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other communication devices with them, making it exceedingly difficult for Niemann to have cheated this way. The specific allegations as to how Niemann could have cheated OTB were and currently are still unknown. However, since none were found prior, any possibility of this has already been eliminated.

One may think the matter would be closed at that, but not quite. Maybe we can't divine anything from a single game, but if a trend could somehow be established then, at the very least Carlsen's suspicions could be validated. That is to say, if the investigation were to propagate into Niemann's history, where statistical analysis would be possible given the amount of data, one might be able to identify suspicious activity. The data is there and Carlsen still holds the public trust, all that needs to be worked out is the how.
Chess.com, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.

Item 2 requires an assessment of move quality which is usually done in centipawns. The centipawn is a unit of measure used in chess as a representation of advantage. One centipawn is equal to one hundredth of a pawn and, although its value plays no formal role in the game, they are used by chess engines and players alike as a means to evaluate the positions of a chess board via centipawn loss. Centipawn loss (CL) is a calculation and numerical score given by a chess engine  (Stockfish, Houdini, Koodo, etc.) as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine.
In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been. A score of 100 means one is losing the equivalent of one pawn per move which would generally be hard to maintain without quickly being mated (unless your opponent misses every chance they get).
